Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-06-20 parallel-tractability-J-v9/optimizedAlgo.tex
fixed the point of \sqsubseteq_*
revision

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\!\tiny rch}}$ where $A(a)\rightarrow B(a)\in P^*$, $\mathcal{O}\models A\sqsubseteq_* B$ and $A(a)\in V$;

@@

\item add \texttt{rch}$(R(a,b)$, $S(a,b))$ to $S_{\textit{\!\tiny rch}}$ where $R(a,b)\rightarrow S(a,b)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq_* S$ and $R(a,b)\in V$;

\item add \texttt{rch}$(R(a,b)$, $S(b,a))$ to $S_{\textit{\!\tiny rch}}$ where $R(a,b)\rightarrow S(b,a)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq_* S^-$ and $R(a,b)\in V$;

@@

\begin{example}\label{exp:prc}
Consider performing Algorithm~$\mathsf{A}_{\text{prc}}$ on the ontology
$\mathcal{O}_{ex_1}$ of Example~\ref{exp:mg}. Note that the individual
set \textbf{IN} is $\{a_1,...,a_k,b\}$, i.e., $k+1$ individuals are involved.
Initially, Algorithm~$\mathsf{A}_{\text{prc}}$ adds
all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) to the
result $\mathcal{G}_{ex_1}$ (\ref{alg4:addFacts}). In the first iteration of
\ref{alg4:updateG}, Algorithm~$\mathsf{A}_{\text{prc}}$ computes $S_{\textit{\!\tiny rch}}$ first.
According to Algorithm~$\mathsf{PRC}$, for each atom of the form
$S(a_i,a_{i-1})$, $2\leq i\leq k$,
and $\forall o\in \mbox{\textbf{IN}}$,
an $\texttt{\emph{rch}}$ relation of the form $\texttt{\emph{rch}}(R(a_{i-1},o),R(a_i,o))$ is added to
$S_{\textit{\!\tiny rch}}$. Since the atom $R(a_1,b)$ has been added to $\mathcal{G}_{ex_1}$,
Algorithm~$\mathsf{A}_{\text{prc}}$ checks that all atoms of the form
$R(a_i,b)$, $2\leq i\leq k$,
have SWD paths and are added to $\mathcal{G}_{ex_1}$ in the first iteration.
In the second iteration of \ref{alg4:updateG}, since all atoms of the
form $R(a_i,b)$, $1\leq i\leq k$,
have been in $\mathcal{G}_{ex_1}$, the $\texttt{\emph{rch}}$ relations of the form $\texttt{\emph{rch}}(A(b),A(a_i))$
are checked with respect to the axiom $\exists R.A\sqsubseteq A$; further all
atoms of the form $A(a_i)$, $1\leq i\leq k$, are finally added to $\mathcal{G}_{ex_1}$
by Algorithm~$\mathsf{A}_{\text{prc}}$.
\end{example}

>>>>

\begin{enumerate}[leftmargin=4ex,label=$\bullet$]
\item add \texttt{rch}$(A(a),B(a))$ to $S_{\textit{\!\tiny rch}}$
  where $A(a)\rightarrow B(a)\in P^*$, $\mathcal{O}\models
  A\sqsubseteq B$ and $A(a)\in V$;

@@

\item add \texttt{rch}$(R(a,b)$, $S(a,b))$ to $S_{\textit{\!\tiny rch}}$ where $R(a,b)\rightarrow S(a,b)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq S$ and $R(a,b)\in V$;

\item add \texttt{rch}$(R(a,b)$, $S(b,a))$ to $S_{\textit{\!\tiny rch}}$ where $R(a,b)\rightarrow S(b,a)\in P^*$,
    $\mathcal{O}\models R\sqsubseteq S^-$ and $R(a,b)\in V$;

@@

\begin{example}\label{exp:prc}
Consider performing Algorithm~$\mathsf{A}_{\text{prc}}$ on the ontology
$\mathcal{O}_{\text{ex}_\ref{exp:mg}}$ of Example~\ref{exp:mg}. Note that the individual
set \textbf{IN} is $\{a_1,...,a_k,b\}$, i.e., $k+1$ individuals are involved.
Initially, Algorithm~$\mathsf{A}_{\text{prc}}$ adds
all the facts ($A(b),R(a_1,b),S(a_2,a_1),...,S(a_{k},a_{k-1})$) to the
result $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ (\ref{alg4:addFacts}). In the first iteration of
\ref{alg4:updateG}, Algorithm~$\mathsf{A}_{\text{prc}}$ computes $S_{\textit{\!\tiny rch}}$ first.
According to Algorithm~$\mathsf{PRC}$, for each atom of the form
$S(a_i,a_{i-1})$, $2\leq i\leq k$,
and $\forall o\in \mbox{\textbf{IN}}$,
an $\texttt{\emph{rch}}$ relation of the form $\texttt{\emph{rch}}(R(a_{i-1},o),R(a_i,o))$ is added to
$S_{\textit{\!\tiny rch}}$. Since the atom $R(a_1,b)$ has been added to $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$,
Algorithm~$\mathsf{A}_{\text{prc}}$ checks that all atoms of the form
$R(a_i,b)$, $2\leq i\leq k$,
have SWD paths and are added to $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ in the first iteration.
In the second iteration of \ref{alg4:updateG}, since all atoms of the
form $R(a_i,b)$, $1\leq i\leq k$,
have been in $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$, the $\texttt{\emph{rch}}$ relations of the form $\texttt{\emph{rch}}(A(b),A(a_i))$
are checked with respect to the axiom $\exists R.A\sqsubseteq A$; further all
atoms of the form $A(a_i)$, $1\leq i\leq k$, are finally added to $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$
by Algorithm~$\mathsf{A}_{\text{prc}}$.
\end{example}

### 2018-06-20 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
revision

For some derived atom $H$, there may exist several rule instantiations where $H$
occurs as a head atom.
This also means that $H$ can be derived in different ways.
The condition in the definition above results in only one way of deriving $H$ being
described by a materialization graph.
% Suppose $\mathcal{G}$ is a materialization graph such that the nodes
% with in-degree 0 are the
% original facts in $\textbf{I}$.
We next give an example of a materialization graph.\\

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{\text{ex}_1}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{\text{ex}_1}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{\text{ex}_1}$,
denoted by $\mathcal{G}_{\text{ex}_1}$.
The nodes with in-degree 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{\text{ex}_1}$ is $k$.
\end{example}

@@

\begin{example}
We consider the datalog program $P_{\text{ex}_1}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{\text{bsc}}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),\ldots,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{\text{ex}_1}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{\text{ex}_1}$ by different processors.
For example a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{\text{ex}_1}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{\text{ex}_1}$.
Algorithm~$\mathsf{A}_{\text{bsc}}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{\text{ex}_1}$ (\ref{alg1:halt}).
\end{example}

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{\text{ex}_1}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{\text{bsc}}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{\text{ex}_1}$.
Let $\mathcal{G}_{\text{ex}_1}$ be some materialization graph corresponding
to some datalog program in $\mathbb{P}_{\text{ex}_1}$. It can be checked that
\texttt{depth}($\mathcal{G}_{\text{ex}_1}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{\text{ex}_1}$ are polynomial in $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{\text{bsc}}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{\text{ex}_1}$.
However, there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{\text{ex}_1}$.
We discuss this in the next part.

@@

In this part, we optimize Algorithm~$\mathsf{A}_{\text{bsc}}$ such that $P_{\text{ex}_1}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{\text{bsc}}$,
we can identify another \texttt{PTD} class.

@@

\begin{example}\label{exp:opt}
We perform Algorithm~$\mathsf{A}_{\text{opt}}$ on the datalog program $P_{\text{ex}_1}$ in Example~\ref{exp:mg}.
Initially, $R(a_1,b)$ is in the materialization graph $\mathcal{G}_{\text{ex}_1}$.
In the first iteration of \ref{alg3:updateG}, all the rule instantiations are in two kinds of forms:
`$R(a_i,b),A(b)\rightarrow A(a_i)$' and `$S(a_i,a_{i-1}),$ $R(a_{i-1},b)\rightarrow R(a_i,b)$', $2\leq i\leq k$,
$S_{\!\textit{\tiny rch}}$ is the set $\{$\texttt{rch}$(R(a_{i-1},b),
R(a_i,b)) \mid 2\leq i\leq k\}
\cup\{$\texttt{rch}$(R(a_i,b), A(a_i)) \mid 1\leq i\leq k\}$.
In the transitive closure of $S_{\!\textit{\tiny rch}}$,
one can check that \texttt{rch}$(R(a_1,b), R(a_i,b)),$
\texttt{rch}$(R(a_1,b), A(a_i))\in S^*_{\!\textit{\tiny rch}}, 2\leq i\leq k$.
Thus, $R(a_i,b)$ and $A(a_i)$, $2\leq i\leq k$, can all be added to
$\mathcal{G}_{\text{ex}_1}$
in the first iteration of \ref{alg3:updateG}.
\end{example}

>>>>

% For some derived atom $H$, there may exist several rule instantiations where $H$
% occurs as a head atom.
% This also means that $H$ can be derived in different ways.
% The condition in the definition above results in only one way of deriving $H$ being
% described by a materialization graph.
% Suppose $\mathcal{G}$ is a materialization graph such that the nodes
% with in-degree 0 are the
% original facts in $\textbf{I}$.
We next give two examples of materialization graphs.\\

\begin{example}\label{exp:matgraph}
  Given a (DL-Lite) ontology $\mathcal{O}_{\text{ex}_\ref{exp:matgraph}}$ where the
  TBox contains the following axioms: $A_1\sqsubseteq B$,
  $A_2\sqsubseteq B$, $A_1\sqsubseteq A_2$; its ABox contains the
  assertions $A_1(a)$, $A_2(a)$ and $B(b)$.  We denote the
  corresponding datalog program of $\mathcal{O}_{\text{ex}_\ref{exp:matgraph}}$ by
  $P_{\text{ex}_\ref{exp:matgraph}}=\langle R, \textbf{I}\rangle$, where $R$ contains
  the rules that are transformed from the above axioms
  ($A_1(x) \to B(x)$, $A_2(x) \to A_1(x)$, $A_1(x) \to A_2(x)$);
  $\textbf{I}$ contains $A_1(a)$, $A_2(a)$ and $B(b)$. The two
  complete materialization graphs of $P_{\text{ex}_\ref{exp:matgraph}}$ are
  shown in Figure~\ref{fig:ex\theexample}.

  Note that despite the rule $A_1(x) \to A_2(x)$, there is no edge
  between the nodes $A_1(a)$ and $A_2(a)$ since $A_1(a)$ and $A_2(a)$
  are original facts from $\textbf{I}$ and, hence, must have an
  in-degree of $0$. Note also that having edges both from $A_1(a)$ to
  $B(a)$ and from $A_2(a)$ to $B(a)$ is not allowed. In this case the
  node $B(a)$ would have the two parents $A_1(a)$ and $A_2(a)$ and,
  hence, there should be a rule
  $A_1(a), A_2(a) \to B(a) \in P_{\text{ex}_\ref{exp:matgraph}}^*$ by the
  definition of materialization graphs, which is not the case. When
  dropping the node $B(a)$ and its incoming edge, we obtain an
  incomplete materialization graph. Note, however, that all facts must
  be contained even in an incomplete materialization graph, e.g.,
  dropping $B(b)$ does not result in a materialization graph.
\end{example}

\begin{figure}[htbp]
\centering
\includegraphics[width=\textwidth]{fig-matgraph}
\caption{The two complete materialization graphs of $\mathcal{O}_{\text{ex}_\ref{exp:matgraph}}$}
\label{fig:ex\theexample}
\end{figure}

The following example serves as a running example within this
section.

\begin{example}\label{exp:mg}
Consider a DHL($\circ$) ontology $\mathcal{O}_{\text{ex}_\ref{exp:mg}}$ where the TBox is
$\{\exists R.A\sqsubseteq A\}$, the RBox is $\{S\circ R\sqsubseteq R\}$ and
the ABox is $\{A(b),R(a_1,b),S(a_i,a_{i-1})\}$ for $2\leq i\leq k$ and
$k$ an integer greater than $2$.
The corresponding datalog program of this ontology is $P_{\text{ex}_\ref{exp:mg}}=\langle R, \textbf{I}\rangle$
where $\textbf{I}$ contains all the assertions in the ABox
and $R$ contains the two rules `$R(x,y),A(y)\rightarrow A(x)$'
and `$S(x,y),R(y,z)\rightarrow R(x,z)$'.
The graph in Figure~\ref{fig:mg} is a materialization graph with respect to $P_{\text{ex}_\ref{exp:mg}}$,
denoted by $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$.
The nodes with in-degree 0 are the original facts in $\textbf{I}$;
each of the other nodes corresponds to a ground instantiation of some rule.
For example, the node $A(a_k)$ corresponds to the ground rule instantiation
`$R(a_k,b),A(b)\rightarrow A(a_k)$'.
The size of this materialization graph is the number of nodes, that is $3k$.
The depth of $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ is $k$.
\end{example}

@@

\begin{example}
We consider the datalog program $P_{\text{ex}_\ref{exp:mg}}$ in Example~\ref{exp:mg} again,
and perform Algorithm~$\mathsf{A}_{\text{bsc}}$ on it.
Initially, all the facts ($A(b),R(a_1,b),S(a_2,a_1),\ldots,S(a_{k},a_{k-1})$) are added to the
result $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ (\ref{alg1:addFacts}).
Then in different iterations of \ref{alg1:updateG}, the remaining nodes are added to
$\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ by different processors.
For example, a processor $p$ is allocated a rule instantiation `$R(a_2,b),A(b)\rightarrow A(a_2)$'.
Then, processor $p$ adds $A(a_2)$ to $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ after it checks that
$A(b)$ and $R(a_2,b)$ are in $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$.
Algorithm~$\mathsf{A}_{\text{bsc}}$ halts when $A(a_k)$ has been added to $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ (\ref{alg1:halt}).
\end{example}

@@

Consider Example~\ref{exp:mg} again. Let the integer $k$ be a variable. We can
get a class of datalog programs, denoted by $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$, where the rule set
is $\{R(x,y),A(y)\rightarrow A(x), S(x,y),R(y,z)\rightarrow R(x,z)\}$ and the fact
set varies according to $k$. The algorithm $\mathsf{A}_{\text{bsc}}^{\psi}$ is restricted in the sense that it cannot even work on the rather simple datalog program class $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$.
Let $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$ be some materialization graph corresponding
to some datalog program in $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$. It can be checked that
\texttt{depth}($\mathcal{G}_{\text{ex}_\ref{exp:mg}}$)$=k$ for some $k$.
This means that the depths of the materialization graphs are linearly bounded by $k$.
On the other hand, the sizes of the datalog programs in $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$ are polynomial in $k$.
Thus, for any $\psi$ that is poly-logarithmically bounded, we can always find a $k$ large
enough such that $\mathsf{A}_{\text{bsc}}^{\psi}$ terminates without constructing a materialization
graph for each datalog program in $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$.
However, there indeed exists an \texttt{NC} algorithm that can handle $\mathbb{P}_{\text{ex}_\ref{exp:mg}}$.
We discuss this in the next part.

@@

In this part, we optimize Algorithm~$\mathsf{A}_{\text{bsc}}$ such that $P_{\text{ex}_\ref{exp:mg}}$
can be handled. Based on the optimized variant of Algorithm~$\mathsf{A}_{\text{bsc}}$,
we can identify another \texttt{PTD} class.

@@

\begin{example}\label{exp:opt}
We perform Algorithm~$\mathsf{A}_{\text{opt}}$ on the datalog program $P_{\text{ex}_\ref{exp:mg}}$ in Example~\ref{exp:mg}.
Initially, $R(a_1,b)$ is in the materialization graph $\mathcal{G}_{\text{ex}_\ref{exp:mg}}$.
In the first iteration of \ref{alg3:updateG}, all the rule instantiations are in two kinds of forms:
`$R(a_i,b),A(b)\rightarrow A(a_i)$' and `$S(a_i,a_{i-1}),$ $R(a_{i-1},b)\rightarrow R(a_i,b)$', $2\leq i\leq k$,
$S_{\!\textit{\tiny rch}}$ is the set $\{$\texttt{rch}$(R(a_{i-1},b),
R(a_i,b)) \mid 2\leq i\leq k\}
\cup\{$\texttt{rch}$(R(a_i,b), A(a_i)) \mid 1\leq i\leq k\}$.
In the transitive closure of $S_{\!\textit{\tiny rch}}$,
one can check that \texttt{rch}$(R(a_1,b), R(a_i,b)),$
\texttt{rch}$(R(a_1,b), A(a_i))\in S^*_{\!\textit{\tiny rch}}, 2\leq i\leq k$.
Thus, $R(a_i,b)$ and $A(a_i)$, $2\leq i\leq k$, can all be added to
$\mathcal{G}_{\text{ex}_\ref{exp:mg}}$
in the first iteration of \ref{alg3:updateG}.
\end{example}
